Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-01 paper.tex
Finish up Sally's comments

Although we emphasise that further work is required to filter out spuriously inferred recombination events, 
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome, 
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often 
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface, 
with the N gene containing fewer potential recombination breakpoints 
than other genes, in agreement with the results of \cite{Turakhia2022-it}. 
Noticeable declines in the number of spanning intervals are also 
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271 
near the beginning of ORF7a and N, respectively), as expected because sites that 
undergo mutation at high rates are more likely to differ between the 
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites 
to appear different between the parents when they are not 
(indeed, there is evidence that 28271

The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
may…).
In other cases, however, the drop in intervals does not appear to 
coincide with hypermutability and may reflect shifts in the actual 
rate of recombination between genes.  
Indeed, template switching is known to occur at hotspots, 
which often involve transcription-regulatory sequences preceding genes 
in SARS-CoV-2~\citep{Yang2021-characterizing}
The rate of recombination may also depend on the relative abundance 
of different subgenomic RNA intermediates that span different genes, 
affecting the availability of templates and influencing the rate of 
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}  
[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here)]

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to being prone to sequencing errors or
for biological reasons such as selection pressure.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site is due to selection.

>>>>

Although we emphasise that further work is required to filter out spuriously inferred recombination events,
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome,
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface,
with the N gene containing fewer potential recombination breakpoints
than other genes, in agreement with the results of \cite{Turakhia2022-it}.
Noticeable declines in the number of spanning intervals are also
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271
near the beginning of ORF7a and N, respectively), as expected because sites that
undergo mutation at high rates are more likely to differ between the
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites
to appear different between the parents when they are not
(see the discussion of potential errors at site 28,271 below).
In other cases, however, the drop in intervals does not appear to
coincide with hypermutability and may reflect shifts in the actual
rate of recombination between genes.
Indeed, template switching is known to occur at hotspots,
which often involve transcription-regulatory sequences preceding genes
in SARS-CoV-2~\citep{Yang2021-characterizing}
The rate of recombination may also depend on the relative abundance
of different subgenomic RNA intermediates that span different genes,
affecting the availability of templates and influencing the rate of
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here)]

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to increased information about parentage
or spurious inferences caused by sequencing errors.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site may be due to selection.
